---
title: Oxygen project using both eICU and MIMIC data
author: Willem van den Boom
date: May 23, 2019
output:
  html_document: default
editor_options:
  chunk_output_type: inline
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup_chunk, results='hide'}
if(!require(bigrquery)) {
  install.packages("bigrquery")
  library("bigrquery")
}

# Project ID from Google Cloud
project_id <- "oxygenators-209612"
options(httr_oauth_cache=FALSE)

# Wrapper for running BigQuery queries.
run_query <- function(query){
    data <- query_exec(query, project=project_id, use_legacy_sql = FALSE, max_pages = Inf)
    return(data)
}

# Library for fitting generalized additive models (GAMs)
library(mgcv)
if(!require(mgcv)) {
  install.packages("mgcv")
  library("mgcv")
}
```


# Read in data from eICU

```{r eval=FALSE}
pat_MIMIC <- run_query('
SELECT * FROM mimiciii_clinical.mimic_final_patient_results
')

pat_eICU <- run_query('
SELECT * FROM eicu.final_patient_results
')
```

```{r include=FALSE, eval=FALSE}
save.image(file = "RData/data_analysis_v4_checkpoint1.RData")
```

```{r loadData, include=FALSE, eval=TRUE}
# Since Google authentication cannot happen when knitting an R Markdown file,
# I load the results from BigQuery from an RData file.
load("RData/data_analysis_v4_checkpoint1.RData")
```


# Data preprocessing



## Age

We recode ethnicity in MIMIC such that its categories match up with eICU.

```{r recode_ethnicity}
pat_eICU$eth <- pat_eICU$ethnicity
pat_eICU$eth[pat_eICU$eth == ""] <- "Other/Unknown"
pat_eICU$eth <- as.factor(pat_eICU$eth)
print(table(pat_eICU$eth))

pat_MIMIC$eth <- pat_MIMIC$ethnicity
pat_MIMIC$eth[grepl("NATIVE", pat_MIMIC$eth)] <- "Native American"
pat_MIMIC$eth[grepl("HISPANIC", pat_MIMIC$eth)] <- "Hispanic"
pat_MIMIC$eth[grepl("WHITE", pat_MIMIC$eth)] <- "Caucasian"
pat_MIMIC$eth[grepl("ASIAN", pat_MIMIC$eth)] <- "Asian"
pat_MIMIC$eth[grepl("AFRICAN AMERICAN", pat_MIMIC$eth)] <- "African American"
pat_MIMIC$eth[grepl("PORTUGUESE", pat_MIMIC$eth)] <- "Caucasian"
pat_MIMIC$eth[grepl("MIDDLE EASTERN", pat_MIMIC$eth)] <- "Caucasian"
pat_MIMIC$eth[!(pat_MIMIC$eth %in% levels(pat_eICU$eth))] <- "Other/Unknown"
pat_MIMIC$eth <- as.factor(pat_MIMIC$eth)

print(table(pat_MIMIC$eth))
```



## Age

Abnormally large ages appear in MIMIC. They are right-censored ages >89. We set them to a missing value.

```{r}
pat_MIMIC$age[pat_MIMIC$age > 100] <- NA
```



## Body mass index

Compute the body mass index of each patient.

```{r}
# Remove heights above 3 meter and below .5 meter.
pat_eICU$height[pat_eICU$height > 300 | pat_eICU$height < 50] <- NA

# Remove weights below 20kg and above 600kg
pat_eICU$weight[pat_eICU$weight < 20 | pat_eICU$weight > 600] <- NA

# Compute the body mass index
pat_eICU$bmi <- pat_eICU$weight / (pat_eICU$height/100)^2

# Remove BMIs below 15 or above 100
pat_eICU$bmi[pat_eICU$bmi < 15 | pat_eICU$bmi > 100] <- NA




# Remove heights above 3 meter and below .5 meter.
pat_MIMIC$height[pat_MIMIC$height > 300 | pat_MIMIC$height < 50] <- NA

# Remove weights below 20kg and above 600kg
pat_MIMIC$weight[pat_MIMIC$weight < 20 | pat_MIMIC$weight > 600] <- NA

# Compute the body mass index
pat_MIMIC$bmi <- pat_MIMIC$weight / (pat_MIMIC$height/100)^2

# Remove BMIs below 15 or above 100
pat_MIMIC$bmi[pat_MIMIC$bmi < 15 | pat_MIMIC$bmi > 100] <- NA
```


## Gender

Transform gender to a factor rather than string
Set genders that are neither male nor female as missing, effectively excluding these from the analysis
```{r}
if(is.character(pat_eICU$gender)) pat_eICU$gender_string <- pat_eICU$gender
summary(as.factor(pat_eICU$gender_string))

pat_eICU$gender <- NA
pat_eICU$gender[pat_eICU$gender_string == "Female"] = "F"
pat_eICU$gender[pat_eICU$gender_string == "Male"] = "M"

pat_eICU$gender <- as.factor(pat_eICU$gender)



if(is.character(pat_MIMIC$gender)) pat_MIMIC$gender_string <- pat_MIMIC$gender
summary(as.factor(pat_MIMIC$gender_string))

pat_MIMIC$gender <- as.factor(pat_MIMIC$gender)
```


## ICU type

Consolidate the various ICU types such that only general, cardiac, and neuro ICU remain.


### From Slack

*Mornin Feng*
The followings are my suggestions:
1/ Just exclude Neuro ICUS
2/ Merge  “Cardiac ICU”
“CCU-CTICU”: Coronary Care Unit / Cardiothoracic ICU
“CTICU”: Cardiothoracic ICU
as CCUs, BUT keep “CSICU”: Cardiac Surgery ICU separately as CSICU
3/ MICU and SICU should not be merged
4/ Med-Surg ICU, KC See what do you think? (edited)


*KC See*
I largely agree with Mornin. The Neuro ICU is unclear what it is - Medical Neuro vs Neurosurgical. For CSICU, I think it is like CTICU. Can we do analysis with and without it, and see if there is a difference? MICU, MSICU, and SICU should be separate. Then if all shows similar results, we can pool everything too

```{r}
pat_eICU$unit_type = as.factor(pat_eICU$unittype)

#levels(pat_eICU$unit_type)[levels(pat_eICU$unit_type) %in% c("Med-Surg ICU", "MICU", "SICU")] = "General ICU"
levels(pat_eICU$unit_type)[levels(pat_eICU$unit_type) %in% c("CCU-CTICU", "CSICU", "CTICU")] = "Cardiac ICU"
```



We do no grouping ICU type for MIMIC since the data come from one hospital. See (https://mimic.physionet.org/mimictables/transfers/) for a list of ICU types.

```{r}
pat_MIMIC$unit_type <- as.factor(pat_MIMIC$unittype)
```



## Determine first stay

For eICU:
We use `hospitalDischargeYear` and `unit_stay_number` to determine whether an ICU stay was a patient's first.

Note that `unit_stay_number` [does not always start at 1](https://github.com/MIT-LCP/eicu-code/issues/16).
```{r firstStay}
pat_eICU$first_stay = NA

for(i in 1:max(pat_eICU$unit_stay_number)) {
  tmp_table = table(pat_eICU$patient_ID[
    pat_eICU$unit_stay_number <= i
    & is.na(pat_eICU$first_stay)
  ])
  
  # An ICU stay that is a patient's only one with `unit_stay_number` <= i
  # must be a first stay.
  tmp_index <- pat_eICU$patient_ID %in% names(tmp_table)[tmp_table == 1]
  
  pat_eICU$first_stay[
     tmp_index
  ] <- FALSE
  
  pat_eICU$first_stay[
    tmp_index
    & pat_eICU$unit_stay_number <= i
  ] <- TRUE
}


# `hospitalDischargeYear` only takes the values 2014 and 2015
unique(pat_eICU$hospitalDischargeYear)

# The remaning NAs in first_stay can only be resolved if a patient
# has only 1 hospital stay with a discharge in 2014
tmp_IDs <- unique(pat_eICU$patient_ID[
  is.na(pat_eICU$first_stay)
  & pat_eICU$hospitalDischargeYear == 2014
])

for(id in tmp_IDs) {
  tmp_index <- which(
    pat_eICU$patient_ID == id &
    pat_eICU$hospitalDischargeYear == 2014
  )
  
  if(length(tmp_index) == 1) {
    pat_eICU$first_stay[pat_eICU$patient_ID == id] <- FALSE
    pat_eICU$first_stay[tmp_index] <- TRUE
  }
}

rm(tmp_table, tmp_index, tmp_IDs, i, id)

summary(pat_eICU$first_stay)
length(unique(pat_eICU$patient_ID))
length(unique(pat_eICU$patient_ID[is.na(pat_eICU$first_stay)]))

summary(pat_MIMIC$first_stay)
```



## Mortality

```{r}
pat_eICU$mortality_in_ICU <- NA
pat_eICU$mortality_in_ICU[pat_eICU$discharge_status_ICU == "Alive"] <- FALSE
pat_eICU$mortality_in_ICU[pat_eICU$discharge_status_ICU == "Expired"] <- TRUE
summary(pat_eICU$mortality_in_ICU)

pat_eICU$mortality_in_Hospt <- NA
pat_eICU$mortality_in_Hospt[pat_eICU$discharge_status_Hospt == "Alive"] <- FALSE
pat_eICU$mortality_in_Hospt[pat_eICU$discharge_status_Hospt == "Expired"] <- TRUE
# If someone expires in the ICU, then they expire in the hospital.
pat_eICU$mortality_in_Hospt[which(pat_eICU$mortality_in_ICU)] <- TRUE
summary(pat_eICU$mortality_in_Hospt)


summary(pat_MIMIC$mortality_in_ICU)
# If someone expires in the ICU, then they expire in the hospital.
pat_MIMIC$mortality_in_Hospt[pat_MIMIC$mortality_in_ICU == 1] <- TRUE
summary(pat_MIMIC$mortality_in_Hospt)
```


## Oxygen saturation measurements

Proportion of SpO2 measurements from 94 to 98
```{r}
pat_eICU$prop <- 1 - pat_eICU$propBelow - pat_eICU$propAbove
pat_MIMIC$prop <- 1 - pat_MIMIC$propBelow - pat_MIMIC$propAbove
```



# Oxygen measurements

We are only interested in those receiving oxygen for at least 48 or `time_limit` hours during their first oxygen therapy session.
We also require them to have at least 24 or `nOxy_limit` blood oxygen saturation (SpO2) measurements during this oxygen therapy session.

```{r countOxygen, include=TRUE, eval=TRUE}
nOxy_limit <- 24
time_limit <- 24*2

# ICU stays that do not have any measurements are currently NA and should be 0.
pat_eICU$nOxy[which(is.na(pat_eICU$nOxy) & !is.na(pat_eICU$vent_duration))] <- 0
pat_MIMIC$nOxy[which(is.na(pat_MIMIC$nOxy) & !is.na(pat_MIMIC$vent_duration))] <- 0
```



# Determine high ventilation

TBD, but left out for now since it is only required for the supplement.
See `eICU_data_analysis_v9.Rmd` for how it can be done.

```{r include=FALSE, eval=FALSE}
save.image(file = "RData/data_analysis_v4_checkpoint_vent2_3.RData")
```



# Subset selection

The following code selects cases of interest while also providing the info required for a flow diagram: That is, how many patients did not meet the inclusion criteria.

```{r}
subset_selection <- function(patient) {

cat("Total number of ICU stays:", nrow(patient))

# We only keep first stays with at least 48 hours of ventilation
tmp <- which(patient$first_stay)
cat("\nNumber of first stays:", length(tmp))
cat("\nNumber of repeat stays:", sum(patient$first_stay == FALSE, na.rm = TRUE))
cat("\nNumber of patients for whom we couldn't determine first stays:", length(unique(patient$patient_ID[is.na(patient$first_stay)])))
patient_subset <- patient[tmp,]



# Ages above 89 are recorded as "> 89" in the eICU data
# and set to approximately 300 in MIMIC.
# The SQL code translates "> 89" to a missing value
# We only consider "adults" (age >= 16).
tmp <- patient_subset$age < 16
sum(patient_subset$age < 16, na.rm = TRUE)
cat("\nPatients younger than 16:", sum(tmp, na.rm = TRUE))
cat("\nPatients with missing or right-censored age:", sum(is.na(tmp)))
tmp[is.na(tmp)] <- TRUE

# `delete` records which cases to delete from patient
delete <- tmp

# Delete those whose gender is unknown or other
tmp <- is.na(patient_subset$gender)
cat("\nPatients with missing or 'other' gender:", sum(tmp))
delete <- delete | tmp


# Delete those with missing BMI
tmp <- is.na(patient_subset$bmi)
cat("\nPatients with missing BMI:", sum(tmp))
delete <- delete | tmp


# Delete those with missing SOFA score
tmp <- is.na(patient_subset$sofatotal)
cat("\nPatients with missing SOFA score:", sum(tmp))
delete <- delete | tmp

# Delete those with missing outcome of hospital mortality
tmp <- is.na(patient_subset$mortality_in_Hospt)
cat("\nPatients with missing outcome of hospital mortality:", sum(tmp))
delete <- delete | tmp



cat("\nPatients selected so far:", sum(!delete))

patient_subset <- patient_subset[!delete,]

tmp_IDs <- patient_subset$icustay_id

tmp <- patient_subset$vent_duration >= time_limit
cat("\nNumber of ICU stays without an oxygen therapy session:", sum(is.na(tmp)))
cat("\nNumber of ICU stays with a first oxygen therapy session of less than 48 hours:", sum(tmp == FALSE, na.rm = TRUE))
tmp <- which(tmp)
cat("\nNumber of first ventilations of at least 48 hours:", length(tmp))
patient_subset <- patient_subset[tmp,]



cat("Number of selected patients so far:", nrow(patient_subset))




# Delete those with fewer than 24 oxygen measurements
delete <- patient_subset$nOxy < nOxy_limit
cat("\nPatients with too few measurements:", sum(delete), "out of", nrow(patient_subset))

# Only keep those who received more oxygen than room air
keep <- patient_subset$supp_oxygen
cat("\nPatients with NA for supp_oxygen:", sum(is.na(keep)))
cat("\nPatients with `supp_oxygen` FALSE:", sum(keep == FALSE, na.rm = TRUE))

patient_subset <- patient_subset[which(!delete & keep),]


### We do not subset by `high_vent_proportion` since this is not recorded for those on supplemental oxygen.
### Tidal volume is also often not recorded for those on noninvasive ventilation.
#table(is.na(patient_subset$high_vent_proportion), patient_subset$oxygen_therapy_type)


cat("\nPatients selected:", nrow(patient_subset))

summary(as.factor(patient_subset$oxygen_therapy_type))
#sum(tmp <- is.na(patient_subset$high_vent_proportion) & patient_subset$oxygen_therapy_type >= 2)


cat("\nCases with supplemental oxygen selected:", sum(patient_subset$oxygen_therapy_type == 1) )

cat("\nCases with mechanical ventilation selected:", sum(patient_subset$oxygen_therapy_type >= 2) )# & !tmp) )

return(list(patient_subset, tmp_IDs))
}

tmp <- subset_selection(pat_eICU)
pat_eICU_subset <- tmp[[1]]
IDs_eICU <- tmp[[2]]

tmp <- subset_selection(pat_MIMIC)
pat_MIMIC_subset <- tmp[[1]]
IDs_MIMIC <- tmp[[2]]

rm(tmp)
```



## Number of unique hospitals

eICU-CRD covers multiple hospitals.
We only keep hospitals with at least 10 cases:
Hospital will be used as a random effect. Having hospitals only appear a few times in the data makes model estimation computationally more challenging.

```{r, include = TRUE, eval = TRUE}
n_hospt <- 10

cat("\n Number of hospitals in current subset:", length(unique(pat_eICU_subset$hospital_id)))

tmp <- table(pat_eICU_subset$hospital_id)
cat("\n Number of hospitals with at least n_hospt patients:", sum(tmp >= n_hospt))

delete <- pat_eICU_subset$hospital_id %in% as.numeric(names(tmp[tmp < n_hospt]))
pat_eICU_subset <- pat_eICU_subset[!delete,]
pat_eICU_subset$hospital_id = as.factor(pat_eICU_subset$hospital_id)

cat("\n Number of patients in selected subset:", nrow(pat_eICU_subset))

rm(tmp, delete)
```



## Demographics

Let us compare the demographics before and after the subset selection.

```{r}
# Variables of which we like summaries
tmp <- c('age', 'gender', 'bmi', 'sofatotal', 'unit_type', 'nOxy', 'mortality_in_Hospt')

summary(pat_eICU[pat_eICU$icustay_id %in% IDs_eICU,tmp])
for(varname in tmp) cat("\n", varname, sd(pat_eICU[pat_eICU$icustay_id %in% IDs_eICU,varname]))

summary(pat_eICU_subset[,tmp])
for(varname in tmp) cat("\n", varname, sd(pat_eICU_subset[,varname]))

summary(pat_eICU_subset$median)
sd(pat_eICU_subset$median)
summary(pat_eICU_subset$prop)
sd(pat_eICU_subset$prop)


summary(pat_MIMIC[pat_MIMIC$icustay_id %in% IDs_MIMIC,tmp])
for(varname in tmp) cat("\n", varname, sd(pat_MIMIC[pat_MIMIC$icustay_id %in% IDs_MIMIC,varname]))

summary(pat_MIMIC_subset[,tmp])
for(varname in tmp) cat("\n", varname, sd(pat_MIMIC_subset[,varname]))

summary(pat_MIMIC_subset$median)
sd(pat_MIMIC_subset$median)
summary(pat_MIMIC_subset$prop)
sd(pat_MIMIC_subset$prop)
```


```{r include=FALSE, eval=FALSE}
save.image(file = "RData/data_analysis_v4_checkpoint_vent3.RData")
```






# Model fitting


Let us first define functions to visualize the GAM fits.
```{r plotFunctions}
plot_width <- 17.2 * 0.393701 # The textwidth of our Word manuscript in inches.
#plot_width <- 4.2126 # The Lancet asks for plots to have a width of 107mm.
font <- "Times" # Text in figures should be Times New Roman according to the Lancet.
point_size <- 10 # Per Lancet instructions


logistic <- function(x) 1/(1+exp(-x))



# Function that adds labels to subfigures, edited from:
# https://logfc.wordpress.com/2017/03/15/adding-figure-labels-a-b-c-in-the-top-left-corner-of-the-plotting-region/
fig_label <- function(label) {
    cex <- 2
    ds <- dev.size("in")
    # xy coordinates of device corners in user coordinates
    x <- grconvertX(c(0, ds[1]), from="in", to="user")
    y <- grconvertY(c(0, ds[2]), from="in", to="user")
    
    # fragment of the device we use to plot
      # account for the fragment of the device that 
      # the figure is using
      fig <- par("fig")
      dx <- (x[2] - x[1])
      dy <- (y[2] - y[1])
      x <- x[1] + dx * fig[1:2]
      y <- y[1] + dy * fig[3:4]
  
  
  sw <- strwidth(label, cex=cex) * 60/100
  sh <- strheight(label, cex=cex) * 60/100
  
  x1 <- x[1] + sw
  
  y1 <- y[2] - sh
  
  old.par <- par(xpd=NA)
  on.exit(par(old.par))
  
  text(x1, y1, label, cex=cex)
}




gam_plotMed <- function(
                    gamfitMed,
                    main = "Median of measurements",
                    xRange = c(92, 100),
                    yRange = c(0, .4),
                    label
) {
print(paste("Number of cases:", summary(gamfitMed)$n))

xlab <- expression("Median oxygen saturation (SpO"[2]*")")

if(colnames(gamfitMed$model)[1] == "mortality_in_Hospt") {
    ylab <- "Probability of hospital mortality"
  } else {
    ylab <- "Probability of ICU mortality"
  }

xName <- colnames(gamfitMed$model)[grep("med", colnames(gamfitMed$model))]

# Color for dotted line
colD <- "Black"

    plot(1, type = 'n', xlim = xRange, ylim = yRange,
         ylab = ylab,
         xlab = xlab, main = main, yaxs = 'i', xaxs = 'i', yaxt = 'n', xaxt = 'n')
    
    att <- pretty(yRange)
if(!isTRUE(all.equal(att, round(att, digits = 2)))) {
  axis(2, at = att, lab = paste0(sprintf('%.1f', att*100), '%'), las = TRUE)
} else axis(2, at = att, lab = paste0(att*100, '%'), las = TRUE)
    
    att <- pretty(xRange)
    axis(1, at = att, lab = paste0(att, '%'), las = TRUE)

    
    eval(parse(text = paste(c('predictionsPlusCI <- predict(gamfitMed, newdata = data.frame(',
                              xName, ' = gamfitMed$model$', xName, ", gender = 'F', age = median(gamfitMed$model$age), bmi = median(gamfitMed$model$bmi),",
                      ifelse(
                        "high_vent_proportion" %in% colnames(gamfitMed$model),
                        "high_vent_proportion = median(gamfitMed$model$high_vent_proportion),",
                        ""
                      ),
                      ifelse(
                        "apsiii" %in% colnames(gamfitMed$model),
                        "apsiii = median(gamfitMed$model$apsiii),",
                        "sofatotal = median(gamfitMed$model$sofatotal),"
                      ),
                      ifelse(
                        "vent_duration" %in% colnames(gamfitMed$model),
                        "vent_duration = median(gamfitMed$model$vent_duration),",
                        ""
                      ),
                      "hospital_id = 264), type = 'link', se.fit = T)"), collapse = "")))

  
  # We have to use the data on which GAM was fit for confidence region as the GAM does not provide standard errors for 'new' input
  eval(parse(text = paste0('xx <- gamfitMed$model$', xName)))
  ord.index <- order(xx)
  xx <- xx[ord.index]
  
  if(gamfitMed$family$link == 'logit') {
    lcl <- logistic(predictionsPlusCI$fit[ord.index] - 1.96*predictionsPlusCI$se.fit[ord.index])
    ucl <- logistic(predictionsPlusCI$fit[ord.index] + 1.96*predictionsPlusCI$se.fit[ord.index])
    
    lines(x = xx, y = lcl, lty = 2, lwd = 2, col = colD)
    lines(x = xx, y = ucl, lty = 2, lwd = 2, col = colD)
    lines(xx, logistic(predictionsPlusCI$fit[ord.index]), lwd = 3)
  } else {
    lcl <- predictionsPlusCI$fit[ord.index] - 1.96*predictionsPlusCI$se.fit[ord.index]
    ucl <- predictionsPlusCI$fit[ord.index] + 1.96*predictionsPlusCI$se.fit[ord.index]
    
    lines(x = xx, y = lcl, lty = 2, lwd = 2)
    lines(x = xx, y = ucl, lty = 2, lwd = 2)
    lines(xx, predictionsPlusCI$fit[ord.index], lwd = 3)
  }
  
  if(!missing(label)) fig_label(label)
}



gam_plotProp <- function(
  gamfitProp,
  main = "Effect of treatment regime",
  yRange = c(0, .3),
  add = FALSE, col = "black",
  label
){
  print(paste("Number of cases:", summary(gamfitProp)$n))
  
  xName <- colnames(gamfitProp$model)[grep("prop", colnames(gamfitProp$model))]
  
  xRange = c(0, max(gamfitProp$model[,xName], na.rm= TRUE))
  
  if(xName == "propBelow") {
    xlab <- expression("Proportion of SpO"[2]*" <94% (black) or >98% (blue)")
    #xlab <- expression("Proportion of  SpO"[2]*" measurements below 94%")
  } else if(xName == "propAbove") {
    xlab <- expression("Proportion of SpO"[2]*" measurements above 98%")
  } else {
    xlab <- expression("Proportion of SpO"[2]*" measurements within 94% to 98%")
  }
  
  if(colnames(gamfitProp$model)[1] == "mortality_in_Hospt") {
    ylab <- "Probability of hospital mortality"
  } else {
    ylab <- "Probability of ICU mortality"
  }
  
  if(!add) {
    plot(1, type = 'n', xlim = xRange, ylim = yRange,
           ylab = ylab,
           xlab = xlab, main = main, yaxs = 'i', xaxs = 'i', yaxt = 'n', xaxt = 'n')
      
      att <- pretty(yRange)
    if(!isTRUE(all.equal(att, round(att, digits = 2)))) {
      axis(2, at = att, lab = paste0(sprintf('%.1f', att*100), '%'), las = TRUE)
    } else axis(2, at = att, lab = paste0(att*100, '%'), las = TRUE)
      
      att <- pretty(xRange)
      axis(1, at = att, lab = paste0(att*100, '%'), las = TRUE)
  
  }
    
    eval(parse(text = paste(c('predictionsPlusCI <- predict(gamfitProp, newdata = data.frame(',
                              xName, ' = gamfitProp$model$', xName, ", gender = 'F', age = median(gamfitProp$model$age), bmi = median(gamfitProp$model$bmi),",
                      ifelse(
                        "high_vent_proportion" %in% colnames(gamfitProp$model),
                        "high_vent_proportion = median(gamfitProp$model$high_vent_proportion),",
                        ""
                      ),
                      ifelse(
                        "apsiii" %in% colnames(gamfitProp$model),
                        "apsiii = median(gamfitProp$model$apsiii),",
                        "sofatotal = median(gamfitProp$model$sofatotal),"
                      ),
                      ifelse(
                        "vent_duration" %in% colnames(gamfitProp$model),
                        "vent_duration = median(gamfitProp$model$vent_duration),",
                        ""
                      ),
                      "hospital_id = 264), type = 'link', se.fit = T)"), collapse = "")))
    
    
  
  # We have to use the data on which GAM was fit for confidence region as the GAM does not provide standard errors for 'new' input
  eval(parse(text = paste0('xx <- gamfitProp$model$', xName)))
  ord.index <- order(xx)
  xx <- xx[ord.index]
  
  if(gamfitProp$family$link == 'logit') {
    lcl <- logistic(predictionsPlusCI$fit[ord.index] - 1.96*predictionsPlusCI$se.fit[ord.index])
    ucl <- logistic(predictionsPlusCI$fit[ord.index] + 1.96*predictionsPlusCI$se.fit[ord.index])
    
    lines(x = xx, y = lcl, lty = 2, lwd = 2, col = col)
    lines(x = xx, y = ucl, lty = 2, lwd = 2, col = col)
    lines(xx, logistic(predictionsPlusCI$fit[ord.index]), lwd = 3, col = col)
  } else {
    lcl <- predictionsPlusCI$fit[ord.index] - 1.96*predictionsPlusCI$se.fit[ord.index]
    ucl <- predictionsPlusCI$fit[ord.index] + 1.96*predictionsPlusCI$se.fit[ord.index]
    
    lines(x = xx, y = lcl, lty = 2, lwd = 2, col = col)
    lines(x = xx, y = ucl, lty = 2, lwd = 2, col = col)
    lines(xx, predictionsPlusCI$fit[ord.index], lwd = 3, col = col)
  }
  
  if(!missing(label)) fig_label(label)
}
```


```{r include=FALSE, eval=FALSE}
save.image("RData/data_analysis_v4_checkpoint2.RData")
```



## Median of measurements

We fit a generalized additive model (GAM) to check for the effect of the median oxygen saturation. GAMs are regression models that allow for nonlinear effects of the predictors.
We add gender and age as predictors to control for them. We also control for hospital but since there are so many hospitals, we add it as a random effect.



```{r}
gamfitMed_eICU <- gamm(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)
```


```{r gamfitMed}
pdf(file = "Figures/med_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

summary(gamfitMed_eICU)
gam_plotMed(gamfitMed_eICU, main = "eICU-CRD", label = "A")

summary(gamfitMed_MIMIC)
gam_plotMed(gamfitMed_MIMIC, main = "MIMIC", label = "B")

dev.off()
```



## Proportion of measurements within range

We now fit a GAM with the proportion of measurements within a range instead of the median of the measurements.

```{r}
gammfitProp_eICU <- gamm(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))
gamfitProp_eICU <- gammfitProp_eICU$gam

gamfitProp_below_eICU <- gamm(mortality_in_Hospt ~ s(propBelow)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitProp_above_eICU <- gamm(mortality_in_Hospt ~ s(propAbove)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam


gamfitProp_MIMIC <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)

gamfitProp_below_MIMIC <- gam(mortality_in_Hospt ~ s(propBelow)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)

gamfitProp_above_MIMIC <- gam(mortality_in_Hospt ~ s(propAbove)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)
```

```{r gamfitProp}
pdf(file = "Figures/prop_v4.pdf", width = plot_width,
    height = 6, pointsize = point_size, family = font)
par(mfcol = c(2,2), cex = 1)

summary(gamfitProp_eICU)
gam_plotProp(gamfitProp_eICU, main = "eICU-CRD", label = "A")

summary(gamfitProp_below_eICU)
gam_plotProp(gamfitProp_below_eICU, main = "", label = "C")

summary(gamfitProp_above_eICU)
gam_plotProp(gamfitProp_above_eICU, main = "", col = "blue", add = TRUE)


summary(gamfitProp_MIMIC)
gam_plotProp(gamfitProp_MIMIC, main = "MIMIC", label = "B")

summary(gamfitProp_below_MIMIC)
gam_plotProp(gamfitProp_below_MIMIC, main = "", label = "D")

summary(gamfitProp_above_MIMIC)
gam_plotProp(gamfitProp_above_MIMIC, main = "", col = "blue", add = TRUE)


dev.off()
```



```{r include=FALSE, eval=FALSE}
save.image("RData/data_analysis_v4_checkpoint3.RData")
```



## Odds ratios

Based on the model fits, we can estimate odds ratios, including 95% confidence intervals.

```{r results='hide'}
if(!require("oddsratio")) {
  install.packages("oddsratio")
  library(oddsratio)
}

computeCI <- function(gam_fit, data) {
  
  if("median" %in% colnames(gam_fit$model)) {
  
  tmp <- or_gam(data = data, model = gam_fit, pred = "median", values = c(96, 100))

  cat(
  "The odds ratio of mortality for a median oxygen saturation at 100% versus 96% is ",
  tmp$oddsratio,
  " (95% CI ",
  tmp$`CI_high (97.5%)`,
  " to ",
  tmp$`CI_low (2.5%)`,
  ").\n",
  sep = ""
  )
  
  tmp <- or_gam(data = data, model = gam_fit, pred = "median", values = c(96, 92))

  cat(
  "The odds ratio of mortality for a median oxygen saturation at 92% versus 96% is ",
  tmp$oddsratio,
  " (95% CI ",
  tmp$`CI_high (97.5%)`,
  " to ",
  tmp$`CI_low (2.5%)`,
  ").",
  sep = ""
  )
  
  } else {
    xName <- colnames(gam_fit$model)[grep("prop", colnames(gam_fit$model))]
    tmp <- or_gam(data = data, model = gam_fit, pred = xName, values = c(.4, .8))

  cat(
  "The odds ratio of mortality for a proportion of measurements within the range at 80% versus 40% is ",
  tmp$oddsratio,
  " (95% CI ",
  tmp$`CI_low (2.5%)`,
  " to ",
  tmp$`CI_high (97.5%)`,
  ").",
  sep = ""
  )
  
  }

}
```

```{r}
computeCI(gamfitMed_eICU, data = pat_eICU_subset)
computeCI(gamfitProp_eICU, data = pat_eICU_subset)
computeCI(gamfitProp_above_eICU, data = pat_eICU_subset)

computeCI(gamfitMed_MIMIC, data = pat_MIMIC_subset)
computeCI(gamfitProp_MIMIC, data = pat_MIMIC_subset)
computeCI(gamfitProp_above_MIMIC, data = pat_MIMIC_subset)
```



Same as above but now without adjusting for covariates

```{r}
tmp <- gam(mortality_in_Hospt ~ s(median), data = pat_eICU_subset, family = binomial)
computeCI(tmp, data = pat_eICU_subset)

tmp <- gam(mortality_in_Hospt ~ s(prop), data = pat_eICU_subset, family = binomial)
computeCI(tmp, data = pat_eICU_subset)

tmp <- gam(mortality_in_Hospt ~ s(propAbove), data = pat_eICU_subset, family = binomial)
computeCI(tmp, data = pat_eICU_subset)


tmp <- gam(mortality_in_Hospt ~ s(median), data = pat_MIMIC_subset, family = binomial)
computeCI(tmp, data = pat_MIMIC_subset)

tmp <- gam(mortality_in_Hospt ~ s(prop), data = pat_MIMIC_subset, family = binomial)
computeCI(tmp, data = pat_MIMIC_subset)

tmp <- gam(mortality_in_Hospt ~ s(propAbove), data = pat_MIMIC_subset, family = binomial)
computeCI(tmp, data = pat_MIMIC_subset)
```


## Supplement

Let us now create additional plots for the supplement.



### ICU mortality instead of hospital mortality

```{r}
gamfitMed_ICU_eICU <- gamm(mortality_in_ICU ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_ICU_MIMIC <- gam(mortality_in_ICU ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)

pdf(file = "Figures/med_ICU_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

summary(gamfitMed_ICU_eICU)
gam_plotMed(gamfitMed_ICU_eICU, main = "eICU-CRD", label = "A")

summary(gamfitMed_ICU_MIMIC)
gam_plotMed(gamfitMed_ICU_MIMIC, main = "MIMIC", label = "B")

dev.off()


gammfitProp_ICU_eICU <- gamm(mortality_in_ICU ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))
gamfitProp_ICU_eICU <- gammfitProp_ICU_eICU$gam

gamfitProp_ICU_MIMIC <- gam(mortality_in_ICU ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)

pdf(file = "Figures/prop_ICU_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

summary(gamfitProp_ICU_eICU)
gam_plotProp(gamfitProp_ICU_eICU, main = "eICU-CRD", label = "A")

summary(gamfitProp_ICU_MIMIC)
gam_plotProp(gamfitProp_ICU_MIMIC, main = "MIMIC", label = "B")


dev.off()
```




### Sensitivity analysis

APACHE score instead of SOFA score as control.

```{r apache_model_fit}
gamfitMed_eICU_tmp <- gamm(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(apsiii)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(apsiii)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)


gamfitProp_eICU_tmp <- gamm(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(apsiii)+s(vent_duration), data = pat_eICU_subset, family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitProp_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(apsiii)+s(vent_duration), data = pat_MIMIC_subset, family = binomial)
```

```{r apache}
pdf(file = "Figures/med_APACHE_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

gam_plotMed(gamfitMed_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotMed(gamfitMed_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()




pdf(file = "Figures/prop_APACHE_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotProp(gamfitProp_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()
```




Only considering those with ventilation session of less than 48 hours.

```{r model_fit_48_reviewer}
gamfitMed_eICU_tmp <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU[pat_eICU$icustay_id %in% IDs_eICU & pat_eICU$supp_oxygen & pat_eICU$vent_duration < 48,], family = binomial)

gamfitMed_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC[pat_MIMIC$icustay_id %in% IDs_MIMIC & pat_MIMIC$supp_oxygen & pat_MIMIC$vent_duration < 48,], family = binomial)

gamfitProp_eICU_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU[pat_eICU$icustay_id %in% IDs_eICU & pat_eICU$supp_oxygen & pat_eICU$vent_duration < 48,], family = binomial)

gamfitProp_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC[pat_MIMIC$icustay_id %in% IDs_MIMIC & pat_MIMIC$supp_oxygen & pat_MIMIC$vent_duration < 48,], family = binomial)
```


```{r limit_48_reviewer}
pdf(file = "Figures/med_48_reviewer_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

gam_plotMed(gamfitMed_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotMed(gamfitMed_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()




pdf(file = "Figures/prop_48_reviewer_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotProp(gamfitProp_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()
```


Only considering the first 24, 48, or 72 hours of a ventilation session when it comes to SpO2.

```{r model_fit_24}
gamfitMed_eICU_tmp <- gamm(mortality_in_Hospt ~ s(median_24)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_24 >= 12,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(median_24)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_24 >= 12,], family = binomial)


gamfitProp_eICU_tmp <- gamm(mortality_in_Hospt ~ s(prop_24)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_24 >= 12,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitProp_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(prop_24)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_24 >= 12,], family = binomial)
```

```{r limit_24}
pdf(file = "Figures/med_24_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

gam_plotMed(gamfitMed_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotMed(gamfitMed_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()




pdf(file = "Figures/prop_24_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotProp(gamfitProp_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()
```



```{r model_fit_48}
gamfitMed_eICU_tmp <- gamm(mortality_in_Hospt ~ s(median_48)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_48 >= 24,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(median_48)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_48 >= 24,], family = binomial)


gamfitProp_eICU_tmp <- gamm(mortality_in_Hospt ~ s(prop_48)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_48 >= 24,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitProp_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(prop_48)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_48 >= 24,], family = binomial)
```

```{r limit_48}
pdf(file = "Figures/med_48_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

gam_plotMed(gamfitMed_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotMed(gamfitMed_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()




pdf(file = "Figures/prop_48_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotProp(gamfitProp_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()
```




```{r model_fit_72}
gamfitMed_eICU_tmp <- gamm(mortality_in_Hospt ~ s(median_72)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_72 >= 36,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(median_72)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_72 >= 36,], family = binomial)


gamfitProp_eICU_tmp <- gamm(mortality_in_Hospt ~ s(prop_72)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$nOxy_72 >= 36,], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitProp_MIMIC_tmp <- gam(mortality_in_Hospt ~ s(prop_72)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$nOxy_72 >= 36,], family = binomial)
```

```{r limit_72}
pdf(file = "Figures/med_72_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

gam_plotMed(gamfitMed_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotMed(gamfitMed_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()




pdf(file = "Figures/prop_72_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)
par(mfcol = c(1,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp, main = "eICU-CRD", label = "A")

gam_plotProp(gamfitProp_MIMIC_tmp, main = "MIMIC", label = "B")

dev.off()
```




### Subgroup analysis


Split by ventilation type

```{r model_fit_vent_type}
gamfitMed_eICU_tmp_1 <- gamm(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$oxygen_therapy_type == 4,], family = binomial, random = list(hospital_id = ~ 1))$gam
gamfitMed_eICU_tmp_2 <- gamm(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$oxygen_therapy_type %in% c(1,3),], family = binomial, random = list(hospital_id = ~ 1))$gam

gamfitMed_MIMIC_tmp_1 <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$oxygen_therapy_type == 4,], family = binomial)
gamfitMed_MIMIC_tmp_2 <- gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$oxygen_therapy_type %in% c(1,3),], family = binomial)


gamfitProp_eICU_tmp_1 <- gamm(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$oxygen_therapy_type == 4,], family = binomial, random = list(hospital_id = ~ 1))$gam
gamfitProp_eICU_tmp_2 <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$oxygen_therapy_type %in% c(1,3),], family = binomial)#, random = list(hospital_id = ~ 1), contol=list(sing.tol=.1, maxIter = 500, msMaxIter = 500, msMaxEval = 500, tolerance = 0.1, msTol = 0.1))$gam # gam instead of gamm because gamm yields singular convergence.

gamfitProp_MIMIC_tmp_1 <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$oxygen_therapy_type == 4,], family = binomial)
gamfitProp_MIMIC_tmp_2 <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$oxygen_therapy_type %in% c(1,3),], family = binomial)
```

```{r vent_type}
pdf(file = "Figures/med_vent_type_v4.pdf", width = plot_width,
    height = 6, pointsize = point_size, family = font)

par(mfcol = c(2,2), cex = 1)

vent_type_2 <- "Supplemental oxygen\nor non-invasive ventilation"

gam_plotMed(gamfitMed_eICU_tmp_1, main = "eICU-CRD\nInvasive ventilation", label = "A")
gam_plotMed(gamfitMed_eICU_tmp_2, main = vent_type_2, label = "C")

gam_plotMed(gamfitMed_MIMIC_tmp_1, main = "MIMIC\nInvasive ventilation", label = "B")
gam_plotMed(gamfitMed_MIMIC_tmp_2, main = vent_type_2, label = "D")

dev.off()


pdf(file = "Figures/prop_vent_type_v4.pdf", width = plot_width,
    height = 6, pointsize = point_size, family = font)
par(mfcol = c(2,2), cex = 1)

gam_plotProp(gamfitProp_eICU_tmp_1, main = "eICU-CRD\nInvasive ventilation", label = "A")
gam_plotProp(gamfitProp_eICU_tmp_2, main = vent_type_2, label = "C")

gam_plotProp(gamfitProp_MIMIC_tmp_1, main = "MIMIC\nInvasive ventilation", label = "B")
gam_plotProp(gamfitProp_MIMIC_tmp_2, main = vent_type_2, label = "D")

dev.off()
```


Look at different ICU types.

```{r ICU_type}
# Matrix with order and "pretty" names for ICU types
mat <- matrix(NA_character_, nrow = 5, ncol = 4)
mat[1,] <- c("Cardiac ICU", "CSRU", "eICU-CRD\nCardiac ICU", "MIMIC\nCardiac surgery recovery unit")
mat[2,] <- c("MICU", "MICU", "Medical ICU", "Medical ICU")
mat[3,] <- c("SICU", "SICU", "Surgical ICU", "Surgical ICU")
mat[4,] <- c("Med-Surg ICU", "CCU", "Medical-surgical ICU", "Coronary care unit")
mat[5,] <- c("Neuro ICU", "TSICU", "Neuro ICU", "Trauma/surgical ICU")



pdf(file = "Figures/med_ICU_type_v4.pdf", width = plot_width,
    height = 12, pointsize = point_size, family = font)

par(mfrow = c(5,2), cex = 1)

for(i in 1:5) {
gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$unit_type == mat[i,1],], family = binomial),
            main = mat[i,3], label = c("A", "C", "E", "G", "I")[i])

gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$unit_type == mat[i,2],], family = binomial),
            main = mat[i,4], label = c("B", "D", "F", "H", "J")[i])

}

dev.off()



pdf(file = "Figures/prop_ICU_type_v4.pdf", width = plot_width,
    height = 12, pointsize = point_size, family = font)

par(mfrow = c(5,2), cex = 1)

for(i in 1:5) {
gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$unit_type == mat[i,1],], family = binomial),
            main = mat[i,3], label = c("A", "C", "E", "G", "I")[i])

gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$unit_type == mat[i,2],], family = binomial),
            main = mat[i,4], label = c("B", "D", "F", "H", "J")[i])

}

dev.off()

```



Split by ethnicity

```{r ethnicity}
# Matrix with order and "pretty" names for ethnicities
n_cat <- 4
mat <- matrix(NA_character_, nrow = n_cat, ncol = 2)
mat[1,] <- c("Caucasian", "Caucasian")
mat[2,] <- c("African American", "African-American")
mat[3,] <- c("Hispanic", "Hispanic")
mat[4,] <- c("Asian", "Asian")



pdf(file = "Figures/med_ethnicity_v4.pdf", width = plot_width,
    height = 12, pointsize = point_size, family = font)

par(mfrow = c(n_cat,2), cex = 1)

for(i in 1:n_cat) {
gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$eth == mat[i,1],], family = binomial),
            main = paste0(ifelse(i == 1, "eICU-CRD\n", ""), mat[i,2]), label = c("A", "C", "E", "G", "I")[i])

gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$eth == mat[i,1],], family = binomial),
            main = paste0(ifelse(i == 1, "MIMIC\n", ""), mat[i,2]), label = c("B", "D", "F", "H", "J")[i])

}

dev.off()



pdf(file = "Figures/prop_ethnicity_v4.pdf", width = plot_width,
    height = 12, pointsize = point_size, family = font)

par(mfrow = c(n_cat,2), cex = 1)

for(i in 1:n_cat) {
gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[pat_eICU_subset$eth == mat[i,1],], family = binomial),
            main = paste0(ifelse(i == 1, "eICU-CRD\n", ""), mat[i,2]), label = c("A", "C", "E", "G", "I")[i])

gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[pat_MIMIC_subset$eth == mat[i,1],], family = binomial),
            main = paste0(ifelse(i == 1, "MIMIC\n", ""), mat[i,2]), label = c("B", "D", "F", "H", "J")[i])

}

dev.off()

```





Split up by whether certain diagnosis was present.

```{r diagnosis}
# Matrix with order and variable/file names for diagnoses
n_cat <- 11
mat <- matrix(NA_character_, nrow = n_cat, ncol = 2)
mat[1,] <- c("has_atrial_fibrillation_disease", "AF")
mat[2,] <- c("has_cancer_disease", "cancer")
mat[3,] <- c("CHF", "CHF")
mat[4,] <- c("CKD", "CKD")
mat[5,] <- c("has_chronic_liver_disease", "CLD")
mat[6,] <- c("has_copd_disease", "COPD")
mat[7,] <- c("has_diabetes_mellitus_disease", "diabetes")
mat[8,] <- c("has_hypertension_disease", "hypertension")
mat[9,] <- c("has_sepsis", "sepsis")
mat[10,] <- c("has_stroke_disease", "stroke")
mat[11,] <- c("has_ischemic_heart_disease", "ischemic heart disease")

for(i in 1:n_cat) {
  print(i)
  print(mat[i,])

pdf(file = paste0("Figures/", mat[i,2], "_v4.pdf"), width = plot_width,
    height = 6, pointsize = point_size, family = font)

par(mfrow = c(2,2), cex = 1)

gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[which(pat_eICU_subset[,mat[i,1]]),], family = binomial),
            main = "eICU-CRD", label = "A", yRange = 0:1)

gam_plotMed(gam(mortality_in_Hospt ~ s(median)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[which(pat_MIMIC_subset[,mat[i,1]]),], family = binomial),
            main = "MIMIC", label = "B", yRange = 0:1)


gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset[which(pat_eICU_subset[,mat[i,1]]),], family = binomial),
            main = "", label = "C", yRange = 0:1)

gam_plotProp(gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_MIMIC_subset[which(pat_MIMIC_subset[,mat[i,1]]),], family = binomial),
            main = "", label = "D", yRange = 0:1)


dev.off()
}
```

Let us also determine the number of cases in the whole dataset.

```{r}
summary(pat_eICU[pat_eICU$icustay_id %in% IDs_eICU, mat[,1]])
summary(pat_MIMIC[pat_MIMIC$icustay_id %in% IDs_MIMIC, mat[,1]])
```



### Histograms of proportions

```{r}
pdf(file = "Figures/hist_v4.pdf", width = plot_width,
    height = 6, pointsize = point_size, family = font)

par(mfcol = c(2,2), cex = 1)

xlab1 <- expression("Proportion of SpO"[2]*" measurements within 94% to 98%")
xlab2 <- expression("Proportion of SpO"[2]*" measurements above 98%")

brks <- seq(from = 0, to = 1, by = .02)
hist(pat_eICU_subset$prop, breaks = brks, main = "eICU-CRD", xlab = xlab1)
fig_label("A")
hist(pat_eICU_subset$propAbove, breaks = brks, xlab = xlab2, main = "")
fig_label("C")

hist(pat_MIMIC_subset$prop, breaks = brks, main = "MIMIC", xlab = xlab1)
fig_label("B")
hist(pat_MIMIC_subset$propAbove, breaks = brks, xlab= xlab2, main = "")
fig_label("D")

dev.off()
```



Let us investigate the correlation between `median` and `vent_duration`.

```{r vent_duration_confounding}
pdf(file = "Figures/vent_duration_confounding_v4.pdf", width = plot_width,
    height = 6, pointsize = point_size, family = font)

par(mfcol = c(2,2), cex = 1)

pat_MIMIC_subset$median_round <- round(pat_MIMIC_subset$median)
pat_MIMIC_subset$median_round[pat_MIMIC_subset$median_round < 92] <- NA
boxplot(vent_duration ~ median_round, data = pat_MIMIC_subset, outline = FALSE, xlab = "Median SpO2 (%)", ylab = "Ventilation duration (hours)", main = "MIMIC")

boxplot(vent_duration ~ round(prop, digits = 1), data = pat_MIMIC_subset, outline = FALSE, xlab = expression("Proportion of SpO"[2]*" measurements within 94% to 98%"), ylab = "Ventilation duration (hours)", main = "MIMIC")


pat_eICU_subset$median_round <- round(pat_eICU_subset$median)
pat_eICU_subset$median_round[pat_eICU_subset$median_round < 92] <- NA
boxplot(vent_duration ~ median_round, data = pat_eICU_subset, outline = FALSE, xlab = "Median SpO2 (%)", ylab = "Ventilation duration (hours)", main = "eICU")

boxplot(vent_duration ~ round(prop, digits = 1), data = pat_eICU_subset, outline = FALSE, xlab = expression("Proportion of SpO"[2]*" measurements within 94% to 98%"), ylab = "Ventilation duration (hours)", main = "eICU")

dev.off()
```



Let us conduct [G computation](doi.org/10.1093/aje/kwq472) to obtain an average treatment effect (ATE), as a causal inference result.

```{r ATE_1}
# Grid of `prop` values
grid <- seq(from = 0, to = 1, by = .1)
grid_size <- length(grid)

ATE <- numeric(grid_size)
tmp_df <- pat_MIMIC_subset

for(i in 1:grid_size) {
  tmp_df$prop <- grid[i]
  ATE[i] <- mean(predict(gamfitProp_MIMIC, newdata = tmp_df, type = "response"))
}

# Number of bootstrap samples to obtain the standard error (SE) of the ATE.
n_boot <- 100

# Vector to store the bootstrap samples
ATE_tmp <- matrix(NA_real_, nrow = grid_size, ncol = n_boot)
ATE_SE <- numeric(grid_size)

set.seed(20190409)
pb <- txtProgressBar(max = n_boot, style = 3)
for(j in 1:n_boot) {
  tmp_df <- pat_MIMIC_subset[sample(1:nrow(pat_MIMIC_subset), replace = TRUE), ]
  gamfit_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = tmp_df, family = binomial)
  
  for(i in 1:grid_size) {
    tmp_df$prop <- grid[i]
    ATE_tmp[i,j] <- mean(predict(gamfit_tmp, newdata = tmp_df, type = "response"))
  }
  
  setTxtProgressBar(pb, j)
}
close(pb)

for(i in 1:grid_size) ATE_SE[i] <- sd(ATE_tmp[i,])

ATE_MIMIC <- ATE
ATE_SE_MIMIC <- ATE_SE
```

```{r ATE_2}
ATE <- numeric(grid_size)
tmp_df <- pat_eICU_subset

gamfit_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = pat_eICU_subset, family = binomial)

for(i in 1:grid_size) {
  tmp_df$prop <- grid[i]
  ATE[i] <- mean(predict(gamfit_tmp, newdata = tmp_df, type = "response"))
}

# Number of bootstrap samples to obtain the standard error (SE) of the ATE.
n_boot <- 100

# Vector to store the bootstrap samples
ATE_tmp <- matrix(NA_real_, nrow = grid_size, ncol = n_boot)
ATE_SE <- numeric(grid_size)

set.seed(20190409)
pb <- txtProgressBar(max = n_boot, style = 3)
for(j in 1:n_boot) {
  tmp_df <- pat_eICU_subset[sample(1:nrow(pat_eICU_subset), replace = TRUE), ]
  gamfit_tmp <- gam(mortality_in_Hospt ~ s(prop)+gender+s(age)+s(bmi)+s(sofatotal)+s(vent_duration), data = tmp_df, family = binomial)
  
  for(i in 1:grid_size) {
    tmp_df$prop <- grid[i]
    ATE_tmp[i,j] <- mean(predict(gamfit_tmp, newdata = tmp_df, type = "response"))
  }
  
  setTxtProgressBar(pb, j)
}
close(pb)

for(i in 1:grid_size) ATE_SE[i] <- sd(ATE_tmp[i,])

ATE_eICU <- ATE
ATE_SE_eICU <- ATE_SE
```

```{r include=FALSE, eval=FALSE}
save.image("RData/data_analysis_v4_checkpoint4.RData")
```

```{r ATE_3}
plot_ATE <- function(
  ATE,
  ATE_SE,
  main = "Effect of treatment regime",
  yRange = c(0, .3),
  add = FALSE, col = "black",
  label
){
  xRange = c(0, 1)
  
 
    xlab <- expression("Proportion of SpO"[2]*" measurements within 94% to 98%")
  
  ylab <- "Probability of hospital mortality" 
  
  if(!add) {
    plot(1, type = 'n', xlim = xRange, ylim = yRange,
           ylab = ylab,
           xlab = xlab, main = main, yaxs = 'i', xaxs = 'i', yaxt = 'n', xaxt = 'n')
      
      att <- pretty(yRange)
    if(!isTRUE(all.equal(att, round(att, digits = 2)))) {
      axis(2, at = att, lab = paste0(sprintf('%.1f', att*100), '%'), las = TRUE)
    } else axis(2, at = att, lab = paste0(att*100, '%'), las = TRUE)
      
      att <- pretty(xRange)
      axis(1, at = att, lab = paste0(att*100, '%'), las = TRUE)
  
  }
    
   
    lcl <- ATE - 1.96*ATE_SE
    ucl <- ATE + 1.96*ATE_SE
    
    lines(x = grid, y = lcl, lty = 2, lwd = 2, col = col)
    lines(x = grid, y = ucl, lty = 2, lwd = 2, col = col)
    lines(grid, ATE, lwd = 3, col = col)
  
  if(!missing(label)) fig_label(label)
}

pdf(file = "Figures/g-comp_v4.pdf", width = plot_width,
    height = 3, pointsize = point_size, family = font)

par(mfcol = c(1,2), cex = 1)

plot_ATE(ATE_eICU, ATE_SE_eICU, main = "eICU-CRD", label = "A")
plot_ATE(ATE_MIMIC, ATE_SE_MIMIC, main = "MIMIC", label = "B")

dev.off()
```
